\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-06-20 doc/incertae_sedis.tex
Update.

\subsection{Ambiguity of split representation for \emph{incertae sedis} taxa}
Equation (\ref{eq:exsetformonea}) introduces a method for encoding a
taxonomy with \emph{incertae sedis} taxa into a set of partial splits.
These partial splits have a unique represention as a matrix of
0-1 characters with missing data indicated by '?'.  Clearly, we will
not be able to recover the identity, or even the number, of characters
that a taxonomist might have used when classifying a set of organisms,
simply by looking at the character matrix that represents a taxonomy
(as Figure \ref{fig:Handling-incertae-sedis-mrp} represents Figure
\ref{fig:Handling-incertae-sedis}b). Nevertheless, one might hope that 
equation (\ref{eq:exsetformonea}) is a unique an uncontroversial
way to encode any taxonomic tree as a set of splits.

@@

The matrix shown in Figure \ref{fig:Placing-one-incertae-MRP}c allows
$V$ to be named in the supertree shown in Figure
\ref{fig:Placing-one-incertae}c because it removes the assumption that
\emph{incertae sedis} taxa can only be placed within siblings.
However, as a side-effect, if taxon $WX$ is found to be
non-monophyletic, the matrix from Figure \ref{fig:Placing-one-incertae-MRP}c
allows $W$ to be placed within $V$.  In general, this approach would
allow incertae sedis children of any taxon $\Gamma_1$ to be placed
arbitrary deep into group $\Gamma_2$ if group $\Gamma_1$ is broken. This
behavior can lead to surprising and non-intuitive results. Additionally, our unpruner currently assumes
that \emph{incertae sedis} taxa 
can only be placed within siblings, and therefore does not know what
to do in this situation.  Thus, it is not clear that either Figure
\ref{fig:Placing-one-incertae-MRP}a or Figure
\ref{fig:Placing-one-incertae-MRP}c comprise a unique, best
encoding of the taxonomy.  Thus, to honor the input taxonomic content
accurately, we may need more taxonomic information than simply a
hierarchy and knowledge of which taxa can float more tipward.

@@

One might therefore hope that running the supertree pipeline on $T_1$,
$T_2$, and $\taxonomy$ simultaneously would yield the same result.
While the topology for the final tree would indeed be the tree shown in 
    Figure \ref{fig:Placing-one-incertae}c, the decision about whether
    or not to name $V$ differs.  As expected,  the definition of $WX$
    in $\taxonomy^\prime$ has been altered to reflect the fact $T_1$
    placed $V$ inside $WX$.  Perhaps unexpected is the fact that
    storing the revised taxonomy with an \emph{incertae sedis}
    annotation and then reprocessing the taxonomy using our rooted
    partial splits approach would also shift the definition of $V$
    from a taxon that must exclude $W$ to one that may include $W$. 
    This results from the fact that equation (\ref{eq:exsetformonea})
    only allows \emph{incertae sedis} taxa to intrude into taxa that
    are their
    sisters in the taxonomy.  Outcomes differ because $W$ is a sister of $V$ in
    $\taxonomy^\prime$ but not $\taxonomy$.  Use of the alternate
    representation of $\taxonomy$ in Figure \ref{fig:Placing-one-incertae-MRP}c
    allows $V$ to be named in Figure \ref{fig:Placing-one-incertae}c,
    which solves the seeming inconsistency.  However, this has some
    surprising side-effects as mentioned above, and is therefore not
    the approach that we present in this paper.

@@

Despite the simplicity of our approach, some complexities and conceptual difficulties
remain.
While the correspondence of nodes and names is a simple one-to-one and onto
relationship without \emph{incertae sedis} taxa, when \emph{incertae sedis} taxa are
introduced, the relationship might be neither one-to-one (Figures \ref{fig:Multiple-names-can},\ref{fig:interdigitation}) nor onto (Figure \ref{fig:One-name-can}).
This necessitates putting more thought into the semantics of taxon names, including
whether names might become synonyms.  This may be considered a ``complexity''.

>>>>

\subsection{Ambiguity of split representation for \emph{incertae sedis} taxa}
Equation (\ref{eq:exsetformonea}) introduces a method for encoding a
taxonomy with \emph{incertae sedis} taxa into a set of partial splits.
These partial splits have a unique represention as a matrix of
0-1 characters with missing data indicated by '?'.  Clearly, we will
not be able to recover the identity, or even the number, of characters
that a taxonomist might have used when classifying a set of organisms,
simply by looking at the character matrix that represents a taxonomy
(as Figure \ref{fig:Handling-incertae-sedis-mrp} represents Figure
\ref{fig:Handling-incertae-sedis}b). Nevertheless, one might hope that 
equation (\ref{eq:exsetformonea}) is a unique and uncontroversial
way to encode any taxonomic tree as a set of splits.

@@

The matrix shown in Figure \ref{fig:Placing-one-incertae-MRP}c allows
$V$ to be named in the supertree shown in Figure
\ref{fig:Placing-one-incertae}c because it removes the assumption that
\emph{incertae sedis} taxa can only be placed within siblings.
However, as a side-effect, if taxon $WX$ is found to be
non-monophyletic, the matrix from Figure \ref{fig:Placing-one-incertae-MRP}c
allows $W$ to be placed within $V$.  In general, this approach would
allow \emph{incertae sedis} children of any taxon $\Gamma_1$ to be placed
arbitrary deep into group $\Gamma_2$ if group $\Gamma_1$ is broken. This
behavior can lead to surprising and non-intuitive results. Additionally, our unpruner currently assumes
that \emph{incertae sedis} taxa 
can only be placed within siblings, and therefore does not know what
to do in this situation.  Thus, it is not clear that either Figure
\ref{fig:Placing-one-incertae-MRP}a or Figure
\ref{fig:Placing-one-incertae-MRP}c comprise a unique, best
encoding of the taxonomy.  Thus, to honor the input taxonomic content
accurately, we may need more taxonomic information than simply a
hierarchy and knowledge of which taxa can float more tipward.

@@

One might therefore hope that running the supertree pipeline on $T_1$,
$T_2$, and $\taxonomy$ simultaneously would yield the same result.
While the topology for the final tree would indeed be the tree shown in 
    Figure \ref{fig:Placing-one-incertae}c, the decision about whether
    or not to name $V$ differs.  As expected,  the definition of $WX$
    in $\taxonomy^\prime$ has been altered to reflect the fact $T_1$
    placed $V$ inside $WX$.  Perhaps unexpected is the fact that
    storing the revised taxonomy with an \emph{incertae sedis}
    annotation and then reprocessing the taxonomy using our rooted
    partial splits approach would also shift the definition of $V$
    from a taxon that must exclude $W$ to one that may include $W$. 
    This results from the fact that equation (\ref{eq:exsetformonea})
    only allows \emph{incertae sedis} taxa to intrude into taxa that
    are their
    sisters in the taxonomy.  Outcomes differ because $W$ is a sister of $V$ in
    $\taxonomy^\prime$ but not $\taxonomy$.  Use of the alternate
    representation of $\taxonomy$ in Figure \ref{fig:Placing-one-incertae-MRP}c
    allows $V$ to be named in Figure \ref{fig:Placing-one-incertae}c,
    which solves the seeming inconsistency.  However, this has some
    surprising side-effects as mentioned above, and is therefore not
    the approach that we focus on here.

@@

Despite the simplicity of our approach, some complexities and conceptual difficulties
remain.
While the correspondence of nodes and names is a simple one-to-one function
without \emph{incertae sedis} taxa, when \emph{incertae sedis} taxa are
introduced, the relationship might be neither one-to-one (Figures
\ref{fig:Multiple-names-can},\ref{fig:interdigitation}) nor a function (Figure \ref{fig:One-name-can}).
This necessitates putting more thought into the semantics of taxon names, including
whether names might become synonyms.  This may be considered a ``complexity''.
